---
title: "Bayes Rules - Chapter 2"
author: "Tural Sadigov"
date: today
format: 
    html:
      theme: 
        light: united
        dark: darkly
      smooth-scroll: true
always_allow_html: true
editor: visual
toc: true
toc-location: right
number-sections: true
cap-location: bottom
fig-cap-location: bottom
tbl-cap-location: top
page-layout: article
code-fold: show
code-summary: "Collapse/show the code below"
code-overflow: scroll
code-line-numbers: true
code-copy: true
execute:
  echo: true
  warning: false
  eval: true
  output: true
  error: false
  cache: true
  freeze: true
  out.width: "100%"
fig-align: center
title-block-banner: true
title-block-banner-color: black
---

# section  2.1

## libraries

```{r}
# Load packages
library(bayesrules)
library(tidyverse)
library(janitor)
```

## data

```{r}
# Import article data
data(fake_news)
```


## EDA

```{r}
fake_news %>% 
  count(type) %>% 
  mutate(prior_pribabilities = n/sum(n)) %>% 
  adorn_totals("row") 
```

Null probability model (probability of an article being FAKE)

$$P(B) = 0.4$$


With 0.5 threshold, we would classify every article as real at the expense of 40% fake ones. 

Let A =  an article has exclamation point in the title, and B = article is fake. Then,

```{r}
fake_news %>% 
  count(type, title_has_excl) %>% 
  group_by(type) %>% 
  mutate(percent = n/sum(n)) %>% 
  adorn_totals("row") 
```

i.e.,

$P(A|B)$ =  P(an article has exclamation point in the title | article is fake) = 26.6%

and 

$P(A|B^c)$ =  P(an article has exclamation point in the title | article is real) = 2.22%

We define likelihood function, $L(. | A)$, when conditioned event is NOT certain, we are evaluating the relative compatibility of data with the uncertain event B. 

$$L(B|A) = P(A|B)$$

On the other hand,

```{r}
fake_news %>% 
  count(title_has_excl, type) %>% 
  group_by(title_has_excl) %>% 
  mutate(percent = n/sum(n)) %>% 
  adorn_totals("row") 

```

i.e., 

Pr(B|A) = 89%, and Pr(B^c|A) = 11%

Note: 

$$P(B|A) + P(B^c|A) = 100%$$ 

but

$$P(A|B) + P(A|B^c) = 28.89% \neq 100%.$$

What if we want to have $P(A) = P(article has excl in its title)$.

```{r}
fake_news %>% 
  count(title_has_excl) %>% 
  mutate(percent = n/sum(n)) %>% 
  adorn_totals("row") 

```

$$P(A) = 12\%, p(A^c) = 88\%$$

Create joint distribution

```{r}
fake_news %>% 
  count(title_has_excl, type) %>% 
  mutate(percent = n/sum(n)) %>% 
  pivot_wider(names_from = type, values_from = percent, id_cols = title_has_excl) %>% 
  adorn_totals('row') %>% 
  adorn_totals('col')
  
```


## Now we include data (features to udpate out model)

We want to calculate posterior probability (probability of the same event after observing more evidence that might affect that event)

$$Pr[B| evidence] = Pr[B|A]$$

```{r}
fake_news %>% 
  filter(title_has_excl == TRUE) %>% 
  count(title_has_excl, type) %>% 
  mutate(percent = n/sum(n)) %>% 
  adorn_totals("row") 
```

# section  2.2

```{r}
data("pop_vs_soda")

pop_vs_soda %>% 
  count(state, sort = TRUE) %>% 
  mutate(state = fct_reorder(state, n)) %>% 
  ggplot(aes(state, n)) +
  geom_col() +
  coord_flip()

pop_vs_soda %>% 
  count(region, sort = TRUE) %>% 
  transmute(region, prior = n/sum(n))

pop_vs_soda %>% 
  group_by(region) %>% 
  count(word_for_cola) %>% 
  filter(word_for_cola == 'pop') %>% 
  ungroup() %>% 
  mutate(percent = n/sum(n)) %>% 
  arrange(percent)


pop_vs_soda %>% 
  mutate(pop_dummy = if_else(word_for_cola == 'pop', TRUE, FALSE)) %>% 
  group_by(pop_dummy) %>% 
  count(region) %>% 
  group_by(region) %>% 
  mutate(percent = n/sum(n)) %>% 
  pivot_wider(names_from = region, values_from = percent, id_cols = pop_dummy) %>% 
  adorn_totals('col') %>% 
  adorn_totals('row')

pop_vs_soda %>% 
  tabyl(pop, region) %>% 
  adorn_percentages("col")

pop_vs_soda %>% 
  tabyl(region) %>% 
  adorn_percentages("col")
```


A = someone uses the word 'pop'
L(M|A) = Pr(A|M) = 0.645 
L(N|A) = Pr(A|N) = 0.273
L(S|A) = Pr(A|S) = 0.0792 
L(W|A) = Pr(A|W) = 0.294 

PRIOR VS Posterior probability
P(S) = 0.241
Pr(S|A) = prior * likelihood/(normalization constant) = 0.241 * 0.0792/0.2826

```{r}
pop_vs_soda %>% 
  mutate(pop_dummy = if_else(word_for_cola == 'pop', TRUE, FALSE)) %>% 
  group_by(pop_dummy) %>% 
  count(region) %>% 
  group_by(region) %>% 
  mutate(percent = n/sum(n)) %>% 
  filter(pop_dummy == TRUE) %>% 
  transmute(region, likelihood = percent) %>% 
  ungroup() %>% 
  bind_cols(pop_vs_soda %>% count(region) %>% transmute(prior = n/sum(n))) %>% 
  mutate(posterior = likelihood*prior/sum(likelihood*prior))
  
```

